---
title: "Calibrating feature weights"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calibrating feature weights}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# The runnable example needs prioritizr and a solver (we use HiGHS here so the
# vignette builds without a commercial licence). Chunks are skipped otherwise.
has_solver <- requireNamespace("prioritizr", quietly = TRUE) &&
  requireNamespace("highs", quietly = TRUE) &&
  requireNamespace("terra", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", eval = has_solver
)
```

## Why calibrate?

ELSA prioritizations maximise a weighted, **additive** utility: the total amount
of each feature secured, summed across planning units and action zones. Because
the objective is additive, spatially extensive features — and features that
co-occur with areas that are valuable to *other* features — are favoured over
rarer or more isolated features when every feature is weighted equally. Some
features end up well represented while others are left far below what they could
achieve.

`elsar_calibrate_weights()` computes a per-feature weight `w_j` that counteracts
this, so features achieve **more even representation relative to their own
single-feature maximum**. It reproduces the ELSA calibration framework exactly;
this vignette shows how to run it and read the result.

## How it works

1. **Single-feature maxima.** Each feature `j` is maximised on its own (weight 1
   for `j`, 0 for the rest) to find its maximum attainable representation
   `a_j^max`.
2. **Baseline.** The problem is solved with equal weights to get each feature's
   representation `a_j` and relative shortfall
   `delta_j = (a_j - a_j^max) / a_j^max`.
3. **Iterate.** Weights are nudged up — most for the worst-represented feature,
   none for the best — the problem is re-solved, and shortfalls recomputed. This
   repeats until the allocation stops changing; the weights giving the most even
   representation (smallest spread of `delta_j`) are returned.

The objective and update rule are unchanged from the original ELSA tool; the
implementation additionally returns the best-seen weights, stops once the
allocation freezes, and excludes zero-total features.

## A worked example

We build a small synthetic stack of four features with different spatial
footprints, two action zones, and per-zone impacts and budgets.

```{r setup}
library(elsar)
library(terra)

set.seed(1)
r0 <- rast(nrows = 24, ncols = 24, xmin = 0, xmax = 24, ymin = 0, ymax = 24,
           crs = "EPSG:3857")
xx <- init(r0, "x")
yy <- init(r0, "y")

features <- c(
  init(r0, runif),                          # broad, diffuse
  (xx < 8) * 1,                             # concentrated strip
  ((xx - 12)^2 + (yy - 12)^2 < 40) * 1,     # central disc
  (yy > 20) * init(r0, runif)               # sparse band
)
names(features) <- c("diffuse", "strip", "disc", "band")

pu_all   <- ifel(!is.na(features[[1]]), 1, NA)
zones_pu <- list(protect = pu_all, restore = pu_all)

impacts <- data.frame(
  feature = names(features),
  protect = c(1.0, 1.0, 1.0, 1.0),
  restore = c(0.6, 0.3, 0.8, 0.5)
)
budgets <- c(protect = 20, restore = 15)  # % of planning-unit area
```

Run the calibration. We use the free HiGHS solver here; in production you would
typically use `solver = "gurobi"`.

```{r calibrate}
cal <- elsar_calibrate_weights(
  features       = features,
  impacts        = impacts,
  zones_pu       = zones_pu,
  planning_units = pu_all,
  budgets        = budgets,
  solver         = "highs",
  verbose        = FALSE
)
```

The calibrated weights:

```{r weights}
cal$weights
```

The representation profile shows how close each feature gets to its single-feature
maximum (`ratio = a / a_max`); calibration lifts the weakest features:

```{r representation}
cal$representation
```

The per-iteration trajectory records how the spread of shortfalls shrinks and when
the allocation froze:

```{r trajectory}
cal$trajectory[, c("iter", "spread_new", "max_abs_util_change", "weight_max")]
```

```{r converged}
cal$converged
cal$iterations
```

## Using the weights

At run time the ELSA tool multiplies these calibration weights by any
user-specified weights, so `w_j = w_j^user * w_j^cal` (with `w_j^user = 1` by
default). Persist `cal$weights` alongside your prepared layers and load them as
the `weight_calibration` for each feature.

## Solvers

`elsar_calibrate_weights()` needs `prioritizr` and a supported solver:

- `solver = "gurobi"` — fastest; requires the commercial `gurobi` R package.
- `solver = "highs"` — free and open-source (`install.packages("highs")`).
- `solver = "default"` — prioritizr's best available solver.

Transient solver or cloud-connection failures are retried automatically
(`solve_retries`).
