---
title: "Global Data Processing Reference"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Global Data Processing Reference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

The `elsar` package provides functions to process various global datasets for
use in ELSA (Essential Life Support Areas) conservation planning. This reference
table summarises the available datasets, their corresponding R functions, and
processing notes.

## Dataset Reference

```{r setup, message=FALSE, echo=FALSE}
library(elsar)
library(readr)
library(dplyr)
library(kableExtra)
```

```{r, echo = FALSE, message=FALSE, warning=FALSE}
data_summary <- read_delim("data_summary.csv", delim = ";", show_col_types = FALSE) %>%
  kable(col.names = c("Dataset", "Dataset Name", "R Function", "Specifics",
                       "Comments", "Tested", "Since")) %>%
  kable_styling("hover", full_width = FALSE, font_size = 10) %>%
  column_spec(1, width = "5cm", bold = TRUE) %>%
  column_spec(2, width = "4cm") %>%
  column_spec(3, width = "4cm") %>%
  column_spec(4, width = "3cm") %>%
  column_spec(5, width = "4cm") %>%
  column_spec(6, width = "2cm") %>%
  column_spec(7, width = "2cm")
data_summary
```

## Notes

- **Tested**: Indicates whether the function has been tested with real data
  (Y = Yes, N = No).
- **Since**: The `elsar` version in which the function was first available.
- Most datasets use `make_normalised_raster()` for standard processing with
  optional parameters (e.g., `inverse = TRUE`).
- Some datasets require specialised functions (e.g., `make_kbas()`,
  `make_mangroves()`, `make_forest_integrity()`).
- Datasets with "NA" as the dataset name do not have a fixed global source file
  and may vary by country or analysis.

## Removed Datasets

The following datasets were previously listed but have been removed or are no
longer supported:

- **Certified Forests** — deprecated, no replacement.
- **Landcover Layers** — replaced by LULC download functions
  (`download_lulc_data()`, `download_lulc_proportions()`).
- **Mountains/Glaciers** — not currently implemented.
- **Productive Managed Forests** — not currently implemented.
- **Urban Extreme Heat** — not currently implemented.
- **Yield Gap** — not currently implemented.
- **Landmark Areas** — not currently implemented.
