---
title: "Global Data Processing Reference"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Global Data Processing Reference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

The `elsar` package provides functions to process various global datasets for use in ELSA (Essential Life Support Areas) conservation planning. This reference table summarizes the available datasets, their corresponding R functions, and processing notes.

## Dataset Reference

```{r setup, message=FALSE, echo=FALSE}
library(elsar)
library(readr)
library(dplyr)
library(kableExtra)
```

```{r, echo = FALSE, message=FALSE, warning=FALSE}
data_summary <- read_delim("data_summary.csv", delim = ";", show_col_types = FALSE) %>%
  kable(col.names = c("Dataset", "Dataset Name", "R Function", "Specifics", "Comments", "Tested")) %>%
  kable_styling("hover", full_width = FALSE, font_size = 10) %>%
  column_spec(1, width = "6cm", bold = TRUE) %>%
  column_spec(2:6, width = "4cm")
data_summary
```

## Notes

- **Tested**: Indicates whether the function has been tested with real data (Y = Yes, N = No)
- Most datasets use `make_normalised_raster()` for standard processing
- Some datasets require specialized functions (e.g., `make_kbas()`, `make_mangroves()`)
- Datasets marked as "NA" or "deprecated" are not currently available
